Constructing a universe for the setoid model

### 2020-06-30 paper.tex
Add reference to agda formalization
Reduce space

@@

We also require indexed W-types, both in $\mType$ and $\mProp$:

\begin{align*}
  W & : (S : \mType) \to (S \to \mType) \to \mType \\
  W_p & : (S : \mType) \to (S \to \mType) \to \mProp
\end{align*}

From $W_p$ we can define propositional truncation:

\begin{align*}
  & \|\_\| : \mType \to \mProp \\
  & |\_| : \{A : \mType \} \to A \to \| A \| \\
  & \textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P
\end{align*}

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:

\begin{align*}
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_P\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
\end{align*}

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We also include a transport combinator
\textsf{transp}.

\begin{align*}
  & \textsf{Id} : \{A : \mType \} \to A \to A \to \mProp \\
  & \textsf{refl} : \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  & \textsf{transp} : \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
\end{align*}

@@

A universe of strict propositions has been recently added to the Agda proof
assistant \cite{agda-prop}, making \emph{most} of $\mlttp$ a subset of Agda,
with the exception of the proof-irrelevant identity type. Most of the universe
constructions presented here have been formalized and proof-checked using Agda,
with the proof-irrelevant identity type and the strong transport rule added via
postulates and rewriting.

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.

\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]

\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
    (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), including function extensionality and a universe of strict
propositions satisfying propositional extensionality \cite{setoid99}. Note that
we don't need neither identity types nor inductive types (W-types) to do this
much.

@@

\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

@@

$\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \\
  \El & = \pi_1
\end{align*}

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes:
%
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
%
in addition to proofs that these are indeed equivalence relations
%
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}

@@

We give an inductive definition of $\setoidU$, mutually with a recursive
definition of everything else. Below we only show the more interesting
definitions of $\setoidU$ and $\El$:
%
\begin{align*}
  & \textsf{data}\ \setoidU : \Set_1 \\
  & \quad \textsf{one} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x' \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
%
\begin{align*}
  & \El\ \textsf{one} :\equiv \textbf{1} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}

@@

We are now ready to give a full definition of the setoid universe:
%
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}

Given $A : \Tm\ \GG\ \Uty$, we define $\EL\ A$ as follows:
%
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}

@@

\begin{align*}
  & \setoidU : \mType_1 \\
  & \setoidU = \Sigma\ (X : \mType)\ (\ad{in-U}\ X) \\
  & \El : \setoidU \to \mType \\
  & \El = \textsf{proj}_1
\end{align*}

@@

\[
A \sim_{\setoidU} B :\equiv \Sigma (R: \El\,A \to \El\,B \to \mProp)(\ad{in-U∼}\ (\pi_2\,A)\ (\pi_2\,B)\ R)
\]

or, using pattern matching:

@@

\begin{align*}
  & \bullet_0 : \textsf{Con}_0 \\
  & \_,_0\_ : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
  & \Pi_0 : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \\
  & \bullet_1 : \textsf{Con}_1\ \bullet_0 \\
  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{align*}

@@

We have described the construction of a universe of setoids in the setoid model
of type theory; this is given in several steps, first as an inductive-recursive
definition, then as an inductive-inductive definition, and finally as an
inductive type. Every encoding is obtained from the previous by adapting known
data type transformation methods in a novel way that accounts for the
peculiarities of our construction.

>>>>

Finally, the mathematical contents of this paper have been formalized in the
proof-assistant Agda (see \cite{agda-code}).

@@

We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W
: (S : \mType) \to (S \to \mType) \to \mType$ and $W_p : (S : \mType) \to (S \to
\mType) \to \mProp$.
%
From $W_p$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
%
\begin{align*}
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_P\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
\end{align*}

Finally, we will assume an identity type in the style of Martin-L\"of's
inductive identity type. The main difference is that our identity type is a
$\mProp$-valued relation. We also include a transport combinator
\textsf{transp}.
%
\begin{align*}
  & \textsf{Id} : \{A : \mType \} \to A \to A \to \mProp \\
  & \textsf{refl} : \{A : \mType \} (a : A) \to \textsf{Id}\ a\ a \\
  & \textsf{transp} : \{A : \mType \} (C : A \to \mType) \{a_0\ a_1 : A\} \to \textsf{Id}\ a_0\ a_1 \to C\ a_0 \to C\ a_1
\end{align*}

@@

A universe of strict propositions has been recently added to the Agda proof
assistant \cite{agda-prop}, making \emph{most} of $\mlttp$ a subset of Agda,
with the exception of the proof-irrelevant identity type. Most of the universe
constructions presented here have been formalized and proof-checked using Agda,
with the proof-irrelevant identity type and the strong transport rule added via
postulates and rewriting. The formalization can be found in \cite{agda-code}.

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
~
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \\
    & \quad \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
    (p : \GG^\sim\ \gamma_0\ \gamma_1) \\
    & \quad \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\
~
\\
We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), including function extensionality and a universe of strict
propositions satisfying propositional extensionality \cite{setoid99}. Note that
we don't need neither identity types nor inductive types (W-types) to do this
much.

@@

\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a)
\end{align*}
\end{minipage}

@@

$\U$ and $\El$ can be given as follows:
\[
\U = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \qquad \El = \pi_1
\]

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations:
%

\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\
~
\\

@@

We give an inductive definition of $\setoidU$, mutually with a recursive
definition of everything else. Below we only show the more interesting
definitions of $\setoidU$ and $\El$:

\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \Set_1 \\
  & \quad \textsf{one} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{one} :\equiv \textbf{1} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$:
%

\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\
~
\\

@@

\begin{minipage}{0.4\textwidth}
\begin{align*}
  & \setoidU : \mType_1 \\
  & \setoidU = \Sigma\ (X : \mType)\ (\ad{in-U}\ X)
\end{align*}
\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \El = \pi_1
\end{align*}
\end{minipage}
\\
~
\\

@@

@@

\begin{minipage}{0.4\textwidth}
\begin{align*}
  & \bullet_0 : \textsf{Con}_0 \\
  & \_,_0\_ : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 : (\GG : \textsf{Con}_0) \to \textsf{Ty}_0 \\
  & \Pi_0 : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \\
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \bullet_1 : \textsf{Con}_1\ \bullet_0 \\
  & \_,_1\_ : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & \qquad \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & \qquad \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & \qquad \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{align*}
\end{minipage}
\\
~
\\

@@

We have described the construction of a universe of setoids in the setoid model
of type theory; this is given in several steps, first as an inductive-recursive
definition, then as an inductive-inductive definition, and finally as an
inductive type. Every encoding is obtained from the previous by adapting known
data type transformation methods in a novel way that accounts for the
peculiarities of our construction. The mathematical contents of this paper have
been formalized in the proof-assistant Agda (see \cite{agda-code}).
